---
title: 'Task 2. Mismanaged plastics: from rivers to oceans'
author: "Anthony Luna"
date: "10/15/2019"
output: html_document
link-citations: true
references:
- id: SCHMIDTinfo
  title: Export of Plastic Debris by Rivers into the Sea
  author: 
  - family: Schmidt,
    given:  Christian
  - family: Krauth
    given:  Tobias
  - family: Wagner
    given:  Stephan
  issued:
    year: 2017
  volume: 51
  number: 21
  pages: 12246-12253
  container-title: Environmental Science \& Technology
  doi: 10.1021/acs.est.7b02368
  URL: 'https://doi.org/10.1021/acs.est.7b02368'
---

```{r setup, include=FALSE}

# Here we set the default such that no warnings or messages will appear in the final document.

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,fig.align = 'center')
```
### Introduction

Globally, 10 rivers account for 88-95% of plastics introduced into the sea [@SCHMIDTinfo]. In this report we will be exploring more regional issues regarding plastic pollution in the seas. We will first look for seas enduring the highest impact of microplastic pollution. We will then focus on the Gulf of Mexico to determine rivers contributing the highest amount of macroplastic pollution.


```{r}
# Here we atrach our required packages

library(tidyverse)
library(janitor)

# Here we load our data into plastic_rivers. We also clean the header names for convienience.

plastic_rivers <- read_csv("plastic_rivers.csv") %>% 
  clean_names()
```


### Part 1. Microplastics Load in Seas

The following graphic provides an overview of seas carry the largest load of microplastics globally.

```{r}
microplastic_sea_load_t15_sum <- plastic_rivers %>% 

  # Here we group by receiving_sea
  
  group_by(receiving_sea) %>% 
  
  # we summarise the data by taking the sum per group
  
  summarise(annual_total_microplastics = sum(microplastic_load_model_2_tons_y_1)) %>% 
  
  # Now we arrange in descending order.
  
  arrange(-annual_total_microplastics) %>% 
  
  # Finally, we take the top 15 elements in this table.
  
  head(15)

# Given this data, we can now visualize the top 15 seas with regard to annual microplastics pollution.

ggplot(microplastic_sea_load_t15_sum,aes(y=annual_total_microplastics, x=receiving_sea )) +
  
  # For this graph, we will use a simple bar chart.
  
  geom_col(fill="blue",size = 100) +
  
  # with this we include a label for each column which has the exact amount of microplastics in tons per year
  
  geom_label(aes(label=scales::comma(annual_total_microplastics)), nudge_y = 6500)+
  
  # Here we set the x - axis to sort by increasing pollution.
  
  scale_x_discrete(limits = rev(microplastic_sea_load_t15_sum$receiving_sea))+
    
  scale_y_continuous(labels = scales::comma, limits = c(0,100000))+
  
   labs(title = "Top 15 Seas by Microplastic Load",
       caption = "Author: Anthony Luna",
       x = "Sea",
       y = expression(paste("Microplastic Load [tons ", y^{-1},"]"))) +
  
  theme_minimal()+
  
  # This centers our title and subtitle
  
  theme(plot.title = element_text(hjust= 0.5),
      plot.subtitle = element_text(hjust= 0.5) )+
  
  # We also will flip the coordinates for aesthetics.
  
  coord_flip()

```


### Part 2. Plastic in the Gulf of Mexico

We are now interested in looking at the Gulf of Mexico in particular. The following graphic shows the top 5 rivers sources for all types of plastic in the Gulf of Mexico.

```{r}

gulf_plastics_t5 <- plastic_rivers %>% 

  # Here we group by receiving_sea
  
  filter(receiving_sea == 'Gulf of Mexico') %>% 
  
  group_by(river) %>% 
  # we summarise the data by taking the sum per group
  
  summarise(annual_total_plastics = sum(macroplastic_load_tons_y_1 + microplastic_load_model_2_tons_y_1)) %>% 
  
  # Now we arrange in descending order.
  
  arrange(-annual_total_plastics) %>% 
  
  # We take the top 5 elements in this table.
  
  head(5) 
  
# Now that we have the data, we can create our plot

ggplot(gulf_plastics_t5,aes(y=annual_total_plastics, x=river )) +

  # We will use a simple bar chart again. This is good practice in terms of providing consistency for the reader  
  
  geom_col(fill = "blue") +
  
  # Here we add labels for clarity
  
  geom_label(aes(label=annual_total_plastics))+
  
  # here we are ordering the rivers from highest to lowest
  
  scale_x_discrete(limits = rev(gulf_plastics_t5$river))+
  
  # Now we flip the coordinates for aesthetics
  
  coord_flip()+
  
  # Here we add a title, caption, and appropriate x and y axes
  
  labs(title = "Top 5 Rivers sources of Plastic Waste in the Gulf of Mexico",
       caption = "Author: Anthony Luna",
       x = "Rivers",
       y = expression(paste("Plastic Load [tons ", y^{-1},"]"))
       )+
  # We use the minimal theme
  
  theme_minimal()+
  
  # This centers our title and subtitle
  
  theme(plot.title = element_text(hjust= 0.5),
      plot.subtitle = element_text(hjust= 0.5))
  

```

### References