---
title: "Lab 3"
author: "Anthony Luna"
date: "10/14/2019"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = FALSE)
```
### Part 1, Attach Packages

```{r}
library(tidyverse)
library(here)
library(janitor)
library(ggridges)
```


### Part 2. Read in the  ouy data (plain text file)

```{r}
sb_buoy <- readr::read_table(here::here("raw_data","sb_buoy_2018.txt"),
                             na = c("99","999","99.00","99.0","999.0")) %>% 
  janitor::clean_names() %>% 
  dplyr::slice(-1) %>% 
  select(number_yy:gst, atmp)
```

### Part 3. Write sb_buoy to a csv file within the "intermediate_data subfolder

```{r}
write_csv(sb_buoy,here("intermediate_data","sb_buoy.csv"))
```

### Part 4. Exploring, and update 'mn' month column to an orered factor withnomth names

First, check it out
```{r, eval=TRUE}
class(sb_buoy$mm)
```

Make class of mm numeric, pull the month names from month.abb
```{r}
sb_buoy_month <- sb_buoy %>% 
  mutate(mm=as.numeric(mm)) %>% 
  mutate(month_name = month.abb[mm])

sb_bouy_fct <- sb_buoy_month %>% 
  mutate(month_name = factor(month_name,levels = month.abb))


```


### Part 5. Exploration and purrr introduction

```{r}
ggplot(sb_bouy_fct, aes(x=month_name, y=atmp)) +
  geom_jitter()
```

Now, we're going to say: for any comlumns that are stored as a character, I would like R to instead understand that they are numeric.


```{r}
sb_buoy_num <- sb_bouy_fct %>% 
  modify_if(is.character,as.numeric)
```



```{r}
ggplot(sb_buoy_num, aes(x=month_name, y=atmp)) +
  geom_violin(aes(fill=month_name)) +
  facet_wrap(~month_name)
```

OK, but I'd like to have those monthly temperatoures plotted separately in facets, but each o them overlayed on top of a population histogram

```{r}
ggplot(sb_buoy_num, aes(x=atmp))+
  geom_histogram(data = transform(sb_buoy_num, month_name=NULL) , fill = "gray80",aes(y=..density..)) +
  geom_histogram(aes(fill=month_name,y=..density..),alpha=.5) +
  facet_wrap(~month_name) +
  theme_light()
```

```{r}
temp_graph <- ggplot(sb_buoy_num, aes(x=atmp,y=month_name)) +
  geom_density_ridges(color = "purple",
                      fill = "orange")+
  scale_x_continuous(limits = c(5,25))+
  scale_y_discrete(limits = rev(levels(sb_buoy_num$month_name))) 

temp_graph


ggsave(here("figures","temp_ridges.png"),
  height = 6,
  width = 6
)

```



Now let's exlore some windspeed:

```{r}
ggplot(sb_buoy_num,aes(wdir,wspd))+
  geom_hex()
```

Now let's see what this looks like if we use a polar coordinate system:

```{r}
ggplot(data = sb_buoy_num, aes(x=wdir, y=wspd)) +
  geom_density2d(size = 0.2,
                 aes(color=month_name),
                 show.legend = FALSE) +
  coord_polar() +
  scale_x_continuous(breaks = c(0,90,180,270),
                     labels = c("N","E","S","W"))+
  facet_wrap(~month_name)+
  theme_minimal()
```

### Part 6. dplyr::case_when() for if-else statements

```{r}
sb_buoy_season <- sb_buoy_num %>% 
  mutate(
    season = case_when(
      month_name %in% c("Mar","Apr","May") ~ "Spring",
      month_name %in% c("Jun","Jul","Aug") ~ "Summer",
      month_name %in% c("Sep","Oct","Nov") ~ "Autumn",
      month_name %in% c("Dec","Jan","Feb") ~ "Winter")
  )
```



